---
title: "AE 04: Randomization test for the slope"
subtitle: "Bikeshare"
author: "Driver: _____, Reporter: ______, Gopher: ______"
date: "Sep 6, 2024"
editor: visual
format: 
  pdf:
    pdf-engine: pdflatex
---

::: callout-important
-   Open [RStudio](https://rstudio.collegeofidaho.edu/) and create a subfolder in your AE folder called "AE-04"

-   Go to the [Canvas](https://cofi.instructure.com/courses/15692/assignments/181485) and locate your `AE 04` assignment to get started.

-   Upload the `ae-04.qmd` and `dcbikeshare.csv` files into the folder you just created. The `.qmd` and PDF responses are due in Canvas no later than Thursday, September 12 at 11:59pm.
:::

```{r load-packages}
#| message: false
library(tidyverse)
library(infer)
library(ggformula)
library(broom)
library(openintro)
library(knitr)
```

## Data

Our dataset contains daily rentals from the Capital Bikeshare in Washington, DC in 2011 and 2012 filtered to only contain the winter months. It was obtained from the `dcbikeshare` data set in the `dsbox` R package.

We will focus on the following variables in the analysis:

-   `count`: total bike rentals
-   `temp_orig`: Temperature in degrees Celsius

```{r glimpse-data}
winter <- read_csv("dcbikeshare.csv") |>
  mutate(season = case_when(
    season == 1 ~ "winter",
    season == 2 ~ "spring",
    season == 3 ~ "summer",
    season == 4 ~ "fall"
  ),
  season = factor(season)) |>
  filter(season == "winter")

glimpse(winter)
```

## Exploratory data analysis

```{r scatterplot}
gf_point(count ~ temp_orig, data = winter, alpha = 0.7) |> 
  gf_labs(
    x = "Temperature (C)",
    y = "Number of Bike Rentals",
  )
```

## Model

```{r fit-model}
model_fit <- lm(count ~ temp_orig, data = winter)

tidy(model_fit) |>
  kable(digits = 2)
```

## Hypothesis test

::: callout-tip
For code chunks with fill-in-the-blank code, change code chunk option to `#| eval: true` once you've filled in the code.
:::

### State the null and alternative hypotheses

\[Add hypotheses in mathematical notation\]

### Generate null distribution using permutation

Fill in the code, then set `eval: true` .

```{r permutation}
#| eval: false

n = 100
set.seed(212)

null_dist <- _____ |>
  specify(______) |>
  hypothesize(null = "independence") |>
  generate(reps = _____, type = "permute") |>
  fit()
```

### Visualize distribution

```{r null-dist}
# Code for histogram of null distribution
```

### Calculate the p-value.

```{r calc-pvalue}
#| eval: false 

# get observed fit 
observed_fit <- winter |>
  specify(count ~ temp_orig) |>
  fit()

# calculate p-value
get_p_value(
  ____,
  obs_stat = ____,
  direction = "two-sided"
)
```

-   What does the warning message mean?

### State conclusion

\[Write your conclusion in the context of the data.\]

::: callout-important
To submit the AE:

-   Render the document to produce the PDF with all of your work from today's class.
-   Upload your PDf and .qmd files to Canvas.
:::
