---
title: "SLR: Model evaluation"
author: "Prof. Eric Friedlander"
date: "2024-09-11"
date-format: "MMM DD, YYYY"
footer: "[ðŸ”— MAT 212 - Fall 2024 -  Schedule](https://mat212fa24.netlify.app/schedule)"
logo: "../images/logo.png"
format: 
  revealjs:
    theme: slides.scss
    multiplex: false
    transition: fade
    slide-number: false
    incremental: false 
    chalkboard: true
html-math-method:
  method: mathjax
  url: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js"
execute:
  freeze: auto
  echo: true
  cache: false
knitr:
  opts_chunk: 
    R.options:      
    width: 200
bibliography: references.bib
---

```{r setup}
#| include: false

library(countdown)

knitr::opts_chunk$set(
  fig.width = 8,
  fig.asp = 0.618,
  fig.retina = 3,
  dpi = 300,
  out.width = "80%",
  fig.align = "center"
)
```

# Questions from last class?

# Model conditions

## Model conditions

1.  **Linearity:** There is a linear relationship between the outcome and predictor variables
2.  **Constant variance:** The variability of the errors is equal for all values of the predictor variable
3.  **Normality:** The errors follow a normal distribution
4.  **Independence:** The errors are independent from each other

## Computational set up

```{r packages}
#| echo: true
#| message: false

# load packages
library(tidyverse)   # for data wrangling and visualization
library(ggformula)   # for plotting using formulas
library(broom)       # for formatting model output
library(scales)      # for pretty axis labels
library(knitr)       # for pretty tables
library(patchwork)   # arrange plots

# HEB Dataset
heb <- read_csv("data/HEBIncome.csv") |> 
  mutate(Avg_Income_K = Avg_Household_Income/1000)

# set default theme and larger font size for ggplot2
ggplot2::theme_set(ggplot2::theme_bw(base_size = 20))
```

## Regression model, revisited

```{r}
#| echo: true
heb_fit <- lm(Number_Organic ~ Avg_Income_K, data = heb)

tidy(heb_fit) |>
  kable(digits = 3)
```

# Model evaluation

## Two statistics {.small}

::: incremental
-   **R-squared**, $R^2$, **Coefficient of Determination** : Percentage of variability in the outcome explained by the regression model (in the context of SLR, the predictor)

    $$
    R^2 = Cor(x,y)^2 = Cor(y, \hat{y})^2
    $$

-   **Root mean square error, RMSE**: A measure of the average error (average difference between observed and predicted values of the outcome)

    $$
    RMSE = \sqrt{\frac{\sum_{i = 1}^n (y_i - \hat{y}_i)^2}{n}}
    $$
:::

. . .

::: question
What indicates a good model fit? Higher or lower $R^2$? Higher or lower RMSE?
:::

## $R^2$

::: incremental
-   Ranges between 0 (terrible predictor) and 1 (perfect predictor)

-   Has no units

-   Calculate with `glance()` or `summary()` using the fitted model:

```{r}
#| echo: true
glance(heb_fit)

glance(heb_fit)$r.squared
```
:::

## Interpreting $R^2$ {.smaller}

```{r}
#| echo: false
heb_fit_rsq <- round(glance(heb_fit)$r.squared * 100, 1)
```

::: poll
ðŸ—³ï¸ **Discussion**

::: midi
::: poll
The $R^2$ of the model for `Number_Organic` from `Average_Income_K` is `r heb_fit_rsq`%. Which of the following is the correct interpretation of this value?
:::

1.  `Avg_Income_K` correctly predicts `r heb_fit_rsq`% of `Number_Organic` in San Anontio HEBs.
2.  `r heb_fit_rsq`% of the variability in `Number_Organic` can be explained by `Avg_Income_K`.
3.  `r heb_fit_rsq`% of the variability in `Avg_Income_K` can be explained by `Number_Organic`.
4.  `r heb_fit_rsq`% of the time `Number_Organic` can be predicted by `Avg_Income_K`.
:::
:::

## RMSE

::: incremental
-   Ranges between 0 (perfect predictor) and infinity (terrible predictor)

-   Same units as the response variable

-   Calculate with `glance()` using the augmented data:

    ```{r}
    #| echo: true

    glance(heb_fit)$sigma
    ```

-   The value of RMSE is not very meaningful on its own, but it's useful for comparing across models (more on this when we get to regression with multiple predictors)
:::

<!-- # Application exercise -->

<!-- ::: appex -->
<!-- ðŸ“‹ [sta210-fa23.netlify.app/ae/ae-06-conditions](https://sta210-fa23.netlify.app/ae/ae-06-conditions.html) -->
<!-- ::: -->