---
title: "SLR: Model evaluation"
author: "Prof. Eric Friedlander"
date: "2024-09-18"
date-format: "MMM DD, YYYY"
footer: "[🔗 MAT 212 - Fall 2024 -  Schedule](https://mat212fa24.netlify.app/schedule)"
logo: "../images/logo.png"
format: 
  revealjs:
    theme: slides.scss
    multiplex: false
    transition: fade
    slide-number: false
    incremental: false 
    chalkboard: true
html-math-method:
  method: mathjax
  url: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js"
execute:
  freeze: auto
  echo: true
  cache: false
knitr:
  opts_chunk: 
    R.options:      
    width: 200
bibliography: references.bib
---

```{r setup}
#| include: false

library(countdown)

knitr::opts_chunk$set(
  fig.width = 8,
  fig.asp = 0.618,
  fig.retina = 3,
  dpi = 300,
  out.width = "80%",
  fig.align = "center"
)
```

# Questions from last class?

# Model conditions

## Model conditions

1.  **Linearity:** There is a linear relationship between the outcome and predictor variables
2.  **Constant variance:** The variability of the errors is equal for all values of the predictor variable
3.  **Normality:** The errors follow a normal distribution
4.  **Independence:** The errors are independent from each other

## Warm-Up: Comparing inferential methods

::: question
-   What are the advantages of using simulation-based inference methods? What are the advantages of using inference methods based on mathematical models?

-   Under what scenario(s) would you prefer to use simulation-based methods? Under what scenario(s) would you prefer to use methods based on mathematical models?
:::

```{r}
#| echo: false
countdown(minutes = 2, font_size = "2em")
```

# Application exercise

::: appex
📋 [AE-06: Model Conditions](https://mat212fa24.netlify.app/ae/ae-06-conditions)
:::

## Computational set up

```{r packages}
#| echo: true
#| message: false

# load packages
library(tidyverse)   # for data wrangling and visualization
library(ggformula)   # for plotting using formulas
library(broom)       # for formatting model output
library(scales)      # for pretty axis labels
library(knitr)       # for pretty tables
library(patchwork)   # arrange plots

# HEB Dataset
heb <- read_csv("data/HEBIncome.csv") |> 
  mutate(Avg_Income_K = Avg_Household_Income/1000)

# set default theme and larger font size for ggplot2
ggplot2::theme_set(ggplot2::theme_bw(base_size = 20))
```

## Regression model, revisited

```{r}
#| echo: true
heb_fit <- lm(Number_Organic ~ Avg_Income_K, data = heb)

tidy(heb_fit) |>
  kable(digits = 3)
```

# Model evaluation

## Partitioning Variability

Let's think about variation:

:::{.incremental}
- DATA = MODEL + ERROR
- $\substack{\text{Variation} \\ \text{in Y}} = \substack{\text{Variation explained} \\ \text{by model}} + \substack{\text{Variation not explained} \\ \text{by model}}$
:::

## Partitioning Variability (ANOVA)

:::{.incremental}
-   $y_i - \bar{y} = (\hat{y}_i - \bar{y}) + (y_i-\hat{y}_i)$
-   Square and sum: $\sum(y_i-\bar{y})^2 = \sum(\hat{y} - \bar{y})^2 + \sum(y-\hat{y})^2$
-   $\substack{\text{Sum of squares} \\ \text{Total}} = \substack{\text{Sum of squares} \\ \text{model}} + \substack{\text{Sum of squares} \\ \text{error}}$
-   $SSTotal = SSModel + SSE$
-   $SST = SSM + SSE$
:::

## ANOVA in R

```{r}
heb_fit |> 
  anova() |> 
  tidy() |> 
  kable()
```

:::{.fragment}
- More on this later in the semester
:::

## Recall: Correlation Coefficient

-   The **correlation coefficient**, $r$, is a number between -1 and +1 that measures how strong the linear relationship between two variables $x$ and $y$ is.

$$
r = \frac{\sum(x_i - \bar{x})(y_i-\bar{y})}{\sqrt{\sum(x_i-\bar{x})^2\sum(y_i-\bar{y})^2}}
= \frac{\sum(x_i - \bar{x})(y_i-\bar{y})}{s_xs_y}
$$


## Two statistics: $R^2$ {.smaller}

::: incremental
-   **R-squared**, $R^2$, **Coefficient of Determination** : Percentage of variability in the outcome explained by the regression model (in the context of SLR, the predictor)
$$
R^2 = Cor(y, \hat{y})^2
$$
    -  Also called **PRE (Percent Reduction in Error)** because:
$$
R^2 = \frac{SSModel}{SSTotal}
$$
:::

 
## Two statistics: RMSE {.smaller}

-   **Root mean square error, RMSE**: A measure of the average error (average difference between observed and predicted values of the outcome)
$$
RMSE = \sqrt{\frac{\sum_{i = 1}^n (y_i - \hat{y}_i)^2}{n}}
$$
    + Sometimes people just case about numerator (SSE) or version without the square-root (MSE)
    + Sometimes the denominator may have $n-1$ instead

. . .

::: question
What indicates a good model fit? Higher or lower $R^2$? Higher or lower RMSE?
:::

## $R^2$

::: incremental
-   Ranges between 0 (terrible predictor) and 1 (perfect predictor)
-   Has no units
-   Calculate with `rsq()` from `yardstick` package using the augmented data:
```{r}
#| echo: true

library(yardstick)
heb_aug <- augment(heb_fit)

rsq(heb_aug, truth = Number_Organic, estimate = .fitted) |> kable()

```
:::

## Interpreting $R^2$ {.smaller}

```{r}
#| echo: false
#| 
heb_fit_rsq <- round(rsq(heb_aug, truth = Number_Organic, estimate = .fitted) |> pull(.estimate) * 100, 1)
```

::: poll
🗳️ **Discussion**

::: midi
::: poll
The $R^2$ of the model for `Number_Organic` from `Average_Income_K` is `r heb_fit_rsq`%. Which of the following is the correct interpretation of this value?
:::

1.  `Avg_Income_K` correctly predicts `r heb_fit_rsq`% of `Number_Organic` in San Antontio HEBs.
2.  `r heb_fit_rsq`% of the variability in `Number_Organic` can be explained by `Avg_Income_K`.
3.  `r heb_fit_rsq`% of the variability in `Avg_Income_K` can be explained by `Number_Organic`.
4.  `r heb_fit_rsq`% of the time `Number_Organic` can be predicted by `Avg_Income_K`.
:::
:::

## Activity

::: appex
In groups, at the board, design a simulation-based procedure for producing a p-value for the following hypothesis test.

- $H_0: R^2 = 0$
- $H_A: R^2 \neq 0$
:::

## RMSE

::: incremental
-   Ranges between 0 (perfect predictor) and infinity (terrible predictor)

-   Same units as the response variable

-   Calculate with  `rmse()` from `yardstick` package using the augmented data:

```{r}
#| echo: true

rmse(heb_aug, truth = Number_Organic, estimate = .fitted)
```

-   The value of RMSE is not very meaningful on its own, but it's useful for comparing across models (more on this and ANOVA when we get to regression with multiple predictors)
:::


# Application exercise

::: appex
📋 [AE-07: Evaluating Models](https://mat212fa24.netlify.app/ae/ae-07-model-eval)
:::