---
title: "Multiple linear regression (MLR)"
subtitle: "Interaction Terms"
author: "Prof. Eric Friedlander"
date: "2024-10-02"
date-format: "MMM DD, YYYY"
footer: "[🔗 MAT 212 - Fall 2024 -  Schedule](https://mat212fa24.netlify.app/schedule)"
logo: "../images/logo.png"
format: 
  revealjs:
    theme: slides.scss
    multiplex: false
    transition: fade
    slide-number: false
    incremental: false 
    chalkboard: true
html-math-method:
  method: mathjax
  url: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js"
execute:
  freeze: auto
  echo: true
  cache: false
knitr:
  opts_chunk: 
    R.options:      
    width: 200
bibliography: references.bib
---

## Announcements

-   Update `eval: false` to `eval: true` at the tops of HW's 03, 04, 05.
-   Resubmit PDFs from 03 and 04
-   Open AE-10

::: appex
📋 [AE 10 - Credit Cards](https://mat212fa24.netlify.app/ae/ae-10-mlr)
:::

## Computational setup

```{r packages}
#| echo: true
#| message: false

# load packages
library(tidyverse)
library(broom)
library(mosaic)
library(ISLR2)
library(patchwork)
library(knitr)
library(kableExtra)
library(scales)

# set default theme and larger font size for ggplot2
ggplot2::theme_set(ggplot2::theme_minimal(base_size = 16))
```

# Last time

## Data: Credit Cards

The data is from the `Credit` data set in the **ISLR2** R package. It is a simulated data set of 400 credit card customers.

```{r}
#| echo: false
Credit |>  glimpse()
```

## Variables

**Features (another name for predictors)**

-   `Income`: Annual income (in 1000's of US dollars)
-   `Rating`: Credit Rating

**Outcome**

-   `Limit`: Credit limit

## Multiple linear regression

The multiple linear regression model assumes

$$
Y|X_1, X_2, \ldots, X_p \sim N(\beta_0 + \beta_1 X_1 + \beta_2 X_2 + \dots + \beta_p X_p, \sigma_{\epsilon}^2)
$$

## Multiple linear regression

At any combination of the predictors, the mean value of the response $Y$, is

$$
\mu_{Y|X_1, \ldots, X_p} = \beta_0 + \beta_1 X_{1} + \beta_2 X_2 + \dots + \beta_p X_p
$$

. . .

Using multiple linear regression, we can estimate the mean response for any combination of predictors

$$
\hat{Y} = \hat{\beta}_0 + \hat{\beta}_1 X_{1} + \hat{\beta}_2 X_2 + \dots + \hat{\beta}_p X_{p}
$$


## Multiple linear regression (MLR) {.smaller}

Based on out analysis goals, we will use a **multiple linear regression** model of the following form

$$
\begin{aligned}\hat{\text{Limit}} ~ = 
\hat{\beta}_0 & + \hat{\beta}_1 \text{Rating} + \hat{\beta}_2 \text{Income}
\end{aligned}
$$

## Model fit

```{r}
#| echo: true
lim_fit <- lm(Limit ~ Rating   + Income,
      data = Credit)

tidy(lim_fit) |>
  kable(digits = 3)
```

## Model equation

$$
\begin{align}\hat{\text{Limit}} =  -532.471 &+14.771 \times \text{Rating}\\
& -0.557 \times \text{Income}
\end{align}
$$

## Visualizing Model

```{r}
#| echo: FALSE
library(plotly)
library(reshape2)
credit_plot <- plot_ly(Credit, 
         x = ~Rating, 
         y = ~Income, 
         z = ~Limit,
         color = I("red"),
         opacity = 1,
         type = "scatter3d", 
         mode = "markers",
         size = 1)

axis_x <- seq(min(Credit$Rating), max(Credit$Rating), length.out = 100)
axis_y <- seq(min(Credit$Income), max(Credit$Income), length.out = 100)

#Sample points
credit_lm_surface <- expand.grid(Rating = axis_x,Income = axis_y,KEEP.OUT.ATTRS = F)
credit_lm_surface$Limit <- predict.lm(lim_fit, newdata = credit_lm_surface)
credit_lm_surface <- acast(credit_lm_surface, Income ~ Rating, value.var = "Limit")

credit_plot |> 
  add_surface(z = credit_lm_surface,
           x = axis_x,
           y = axis_y,
           opacity = 0.8,
           inherit = FALSE,
           contours = list(z = list(show = TRUE, width = 2,
                                       highlightcolor = "#ff0000", 
                                       project = list(z = TRUE)),
                              coloring = "heatmap"))
```


# Interaction terms

## Interaction terms {.smaller}

-   Sometimes the relationship between a predictor variable and the response depends on the value of another predictor variable.
-   This is an **interaction effect**.
-   To account for this, we can include **interaction terms** in the model.
-   We want a model of the form:

$$
\begin{aligned}\hat{\text{Limit}} ~ = 
\hat{\beta}_0 & + \hat{\beta}_1 \text{Rating} + \hat{\beta}_2 \text{Income} +
\hat{\beta}_3\text{Rating}\times\text{Income}
\end{aligned}
$$

## Interpreting interaction terms

-   What the interaction means: The effect of annual income on the credit limit depends on the borrowers credit rating

## Visualizing Interaction Model Exaggerated

```{r}
#| echo: FALSE

lim_fit_int <- lm(Limit ~ Rating  + Income + Rating * Income,
      data = Credit)

lim_fit_int$coefficients[3] <- 200

#Sample points
credit_lm_surface <- expand.grid(Rating = axis_x,Income = axis_y,KEEP.OUT.ATTRS = F)
credit_lm_surface$Limit <- predict.lm(lim_fit_int, newdata = credit_lm_surface)
credit_lm_surface <- acast(credit_lm_surface, Income ~ Rating, value.var = "Limit")

credit_plot |> 
  add_surface(z = credit_lm_surface,
           x = axis_x,
           y = axis_y,
           opacity = 0.8,
           inherit = FALSE,
           contours = list(z = list(show = TRUE, width = 2,
                                       highlightcolor = "#ff0000", 
                                       project = list(z = TRUE)),
                              coloring = "heatmap"))
```

## Visualizing Interaction Model

```{r}
#| echo: FALSE

lim_fit_int <- lm(Limit ~ Rating  + Income + Rating * Income,
      data = Credit)

#Sample points
credit_lm_surface <- expand.grid(Rating = axis_x,Income = axis_y,KEEP.OUT.ATTRS = F)
credit_lm_surface$Limit <- predict.lm(lim_fit_int, newdata = credit_lm_surface)
credit_lm_surface <- acast(credit_lm_surface, Income ~ Rating, value.var = "Limit")

credit_plot |> 
  add_surface(z = credit_lm_surface,
           x = axis_x,
           y = axis_y,
           opacity = 0.8,
           inherit = FALSE,
           contours = list(z = list(show = TRUE, width = 2,
                                       highlightcolor = "#ff0000", 
                                       project = list(z = TRUE)),
                              coloring = "heatmap"))
```

## Application Exercise

::: appex
Complete Activity
:::

## Model Fit {.smaller}

```{r}
#| echo: FALSE
lm(Limit ~ Income + Rating + Income * Rating, data = Credit) |> 
  tidy() |> 
  kable(digits = 5)
```

$$
\begin{aligned}\hat{\text{Limit}} ~ = 
&  -579.85561 + 14.87125~\text{Rating} + 1.81240~\text{Income}\\
& \qquad- 0.00221~\text{Rating}\times\text{Income}
\end{aligned}
$$

## Interpreting the interaction term

-  For a fixed `Rating` the slope of `Income` is $(1.81240 - 0.0021\times\text{Rating})$
-  For a fixed `Income` the slope of `Rating` is $(14.87125 - 0.0021\times\text{Income})$



## What's actually happening: {.smaller}

```{r}
#| echo: TRUE
Credit_int <- Credit |> 
  mutate(Interaction = Income * Rating) 

Credit_int |> 
  select(Limit, Income, Rating, Interaction) |> 
  head() |> 
  kable()
```

## What's actually happening:

```{r}
lm(Limit ~ Income + Rating + Interaction, data = Credit_int) |> 
  tidy() |> 
  kable(digits = 5)
```


# Wrap up

## Recap

-   Introduced multiple linear regression

-   Interpreted coefficients in the multiple linear regression model

-   Calculated predictions and associated intervals for multiple linear regression models

-   Used interaction terms
