---
title: "Model Selection"
author: "Prof. Eric Friedlander"
date: "2024-10-28"
date-format: "MMM DD, YYYY"
footer: "[🔗 MAT 212 - Fall 2024 -  Schedule](https://mat212fa24.netlify.app/schedule)"
logo: "../images/logo.png"
format: 
  revealjs:
    theme: slides.scss
    multiplex: false
    transition: fade
    slide-number: false
    incremental: false 
    chalkboard: true
html-math-method:
  method: mathjax
  url: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js"
execute:
  freeze: auto
  echo: true
  cache: false
knitr:
  opts_chunk: 
    R.options:      
    width: 200
bibliography: references.bib
---

```{r}
#| include: false

# figure options
knitr::opts_chunk$set(
  fig.width = 10, fig.asp = 0.618,
  fig.retina = 3, dpi = 300, fig.align = "center"
)
```

## Announcements

-   Project: EDA Due Wednesday
-   Oral R Quiz

::: appex
📋 [AE 16 - Model Selection](https://mat212fa24.netlify.app/ae/ae-15-comparison2)

- Open up AE 16
:::


## Topics

::: nonincremental
-   Model Selection
-   ANOVA, revisited
:::


## Computational setup

```{r}
#| echo: true

# load packages
library(tidyverse)
library(broom)
library(yardstick)
library(ggformula)
library(GGally)
library(tidymodels)
library(patchwork)
library(knitr)
library(janitor)
library(kableExtra)

# set default theme and larger font size for ggplot2
ggplot2::theme_set(ggplot2::theme_bw(base_size = 20))
```

# Introduction

## Data: Restaurant tips

Which variables help us predict the amount customers tip at a restaurant?

```{r}
#| echo: false
#| message: false
tips <- read_csv("data/tip-data.csv") |>
  drop_na(Party) |> 
  select(-`Tip Percentage`, -`W/Tip`)
```

```{r}
#| echo: false
tips |>
  head() |> 
  kable()
```

## Variables

**Predictors**:

::: nonincremental
-   `Party`: Number of people in the party
-   `Meal`: Time of day (`Lunch`, `Dinner`, `Late Night`)
-   `Age`: Age category of person paying the bill (`Yadult`, `Middle`, `SenCit`)
-   `Day`: Day of the week
- 	`Payment`: the type of payment used	
-   `GiftCard`: whether a giftcard was used	
-   `Comps`: Whether any food was comped\
-   `Alcohol`: Whether any alcohol was ordered
-   `Bday`: whether it was someones birthday	
-   `Bill`: the size of the bill
:::

**Outcome**: `Tip`: Amount of tip

# Model Selection

## Model Selection

-   So far: We've come up with a variety of metrics which help us compare different models
-   How do we choose the models to compare in the first place?
-   Today: Best subset, forward selection, and backward selection


## Recap

-   Comparing models with
    -   $R^2$ vs. $R^2_{Adj}$
    -   AIC and BIC
-   Occam's razor and parsimony

-   Complete Exercise 5.

