---
title: "Model Selection"
author: "Prof. Eric Friedlander"
date: "2024-10-28"
date-format: "MMM DD, YYYY"
footer: "[🔗 MAT 212 - Fall 2024 -  Schedule](https://mat212fa24.netlify.app/schedule)"
logo: "../images/logo.png"
format: 
  revealjs:
    theme: slides.scss
    multiplex: false
    transition: fade
    slide-number: false
    incremental: false 
    chalkboard: true
html-math-method:
  method: mathjax
  url: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js"
execute:
  freeze: auto
  echo: true
  cache: false
knitr:
  opts_chunk: 
    R.options:      
    width: 200
bibliography: references.bib
---

```{r}
#| include: false

# figure options
knitr::opts_chunk$set(
  fig.width = 10, fig.asp = 0.618,
  fig.retina = 3, dpi = 300, fig.align = "center"
)
```

## Announcements

-   Project: EDA Due Wednesday
-   Oral R Quiz

::: appex
📋 [AE 16 - Model Selection](https://mat212fa24.netlify.app/ae/ae-16-modelselection)

- Open up AE 16 and Complete Exercise 0
:::


## Topics

::: nonincremental
-   ANOVA, revisited
-   Model Selection
:::


## Computational setup

```{r}
#| echo: true

# load packages
library(tidyverse)
library(broom)
library(yardstick)
library(ggformula)
library(GGally)
library(tidymodels)
library(patchwork)
library(knitr)
library(janitor)
library(kableExtra)

# set default theme and larger font size for ggplot2
ggplot2::theme_set(ggplot2::theme_bw(base_size = 20))
```

# Introduction

## Data: Restaurant tips

Which variables help us predict the amount customers tip at a restaurant?

```{r}
#| echo: false
#| message: false
tips <- read_csv("data/tip-data.csv") |>
  drop_na(Party) |> 
  select(-`Tip Percentage`, -`W/Tip`)
```

```{r}
#| echo: false
tips |>
  head() |> 
  kable()
```

## Variables {.smaller}

**Predictors**:

::: nonincremental
-   `Party`: Number of people in the party
-   `Meal`: Time of day (`Lunch`, `Dinner`, `Late Night`)
-   `Age`: Age category of person paying the bill (`Yadult`, `Middle`, `SenCit`)
-   `Day`: Day of the week
- 	`Payment`: the type of payment used	
-   `GiftCard`: whether a giftcard was used	
-   `Comps`: Whether any food was comped\
-   `Alcohol`: Whether any alcohol was ordered
-   `Bday`: whether it was someones birthday	
-   `Bill`: the size of the bill
:::

**Outcome**: `Tip`: Amount of tip

# Comparing sets of predictors

## Nested Models {.smaller}

-   We say one model is **nested** inside another model if all of its **TERMS** are present in the other model

. . .

-   Consider three different models:
    -   **Model 1:** $Y = \beta_0 + \beta_1X_1 + \beta_2X_2 + \beta_3X_3 + \epsilon$
    -   **Model 2:** $Y = \beta_0 + \beta_1X_1 + \beta_2X_2 + \epsilon$
    -   **Model 3:** $Y = \beta_0 + \beta_1X_1 + \beta_2X_2 + \beta_3X_1X_2 + \epsilon$
  
. . .

-   **Model 2** is nested inside both **Model 1** and **Model 3**. 
-   Why isn't **Model 3** nested in **Model 1**?
-   Smaller model is called the **Reduced Model**
-   Larger model is called the **Full Model** (be careful, this term depends on context)

. . .

- Complete Exercises 1-2.

## Recall: ANOVA, F-Test {.smaller}

**Hypotheses:** 

$H_0: \beta_1 = \beta_2 = \cdots = \beta_k = 0$ vs. $H_A:$ at least one $\beta_i \neq 0$ 

**Test statistic:** F-statistic

$$
F = \frac{MSModel}{MSE} = \frac{SSModel/k}{SSE/(n-k-1)} \\
$$

**p-value:** Probability of observing a test statistic at least as extreme (in the direction of the alternative hypothesis) from the null value as the one observed

$$
\text{p-value} = P(F > \text{test statistic}),
$$

calculated from an $F$ distribution with $k$ and $n - k - 1$ degrees of freedom.

## Nested F-Test {.smaller}

Suppose $k_1$ is the number of $\beta$'s in the nested model and $k$ is the full number of predictors in the larger model. I.e. $\beta_{k1+1},\ldots, \beta_{k}$ are the new $\beta$'s

**Hypotheses:** 

$H_0: \beta_{k_1+1} = \beta_{k_1+2} = \cdots = \beta_k = 0$ vs. $H_A:$ at least one $\beta_i \neq 0$ for $i>k_1+1$ 

**Test statistic:** F-statistic

$$
F = \frac{(SSModel_{full} - SSModel_{reduced})/(k-k_1)}{SSE_{full}/(n-k-1)} \\
$$

**p-value:** Probability of observing a test statistic at least as extreme (in the direction of the alternative hypothesis) from the null value as the one observed

$$
\text{p-value} = P(F > \text{test statistic}),
$$

calculated from an $F$ distribution with $k-k_1$ (the number of predictors being tested) and $n - k - 1$ degrees of freedom.

. . .

Note: Same as regular F-test if reduced model is just $Y= \beta_0$.

## Nested F-Test in R {.smaller}

```{r}
tip_fit_1 <- lm(Tip ~ Party + Age + Meal, data = tips)
tip_fit_2 <- lm(Tip ~ Party + Age + Meal + Day, data = tips)

anova(tip_fit_1, tip_fit_2) |> # Enter reduced model first
  tidy() |> 
  kable()
```

$$
F = \frac{(SSModel_{full} - SSModel_{reduced})/(k-k_1)}{SSE_{full}/(n-k-1)}
= \frac{15.59778/5}{6073815/158}
= 0.8114993
$$

Let's interpret this together.

. . .

Complete Exercise 3.


# Model Selection

## Model Selection

-   So far: We've come up with a variety of metrics and tests which help us compare different models
-   How do we choose the models to compare in the first place?
-   Today: Best subset, forward selection, and backward selection

## AIC, BIC, Mallows' $C_p$

Estimators of prediction error and *relative* quality of models:

**Akaike's Information Criterion (AIC)**: $$AIC = n\log(SS_\text{Error}) - n \log(n) + 2(p+1)$$ <br>

**Schwarz's Bayesian Information Criterion (BIC)**: $$BIC = n\log(SS_\text{Error}) - n\log(n) + log(n)\times(p+1)$$

. . .

**Mallows' $C_p$**: $$C_p = \frac{SSE_{p}}{MSE_{full model}} - n + 2(p+1)$$

## Best Subset Selection

-   Computers are great now!
-   Frequently feasible to try out EVERY combination of predictors if you total number of possible predictors is not too high.

## Best Subset Selection in R {.smaller}

```{r}
library(leaps)

regsubsets(Tip ~ ., data = tips) |> 
  tidy() |> 
  kable()
```

Shows you "best" model for every model size.

## Backward Elimination {.smaller}

Different model selection technique:

1. Start by fitting the full model (the model that includes all terms under consideration).
2. Identify the term with the largest p-value.
    a. If p-value is large (say, greater than 5%), eliminate that term to produce a smaller model. Fit that model and return to the start of Step 2.
    b. If p-value is small (less than 5%), stop since all of the predictors in the model are "significant."

Note: this can be altered to work with other criterea (e.g. AIC) instead of p-values. This is actually what `regsubsets` does.


## Forward Selection {.smaller}

1. Start with a model with no predictors and find the best single predictor (the largest correlation with the response gives the biggest initial).
2. Add the new predictor to the model, run the regression, and find its individual p-value:
    a. If p-value is small (say, less than 5%), add predictor which would produce the most benefit (biggest increase in $R^2$) when added to the existing model.
    b. If the p-value is large (over 5%), stop and discard this predictor. At this point, no (unused) predictor should be significant when added to the model and we are done.

## Stepwise Selection {.smaller}

Forward, stepwise selection

1. Start with a model with no predictors and find the best single predictor (the largest correlation with the response gives the biggest initial).
2. Add the new predictor to the model, run the regression, and find its individual p-value:
    a. If p-value is small (say, less than 5%), **run backward elimination**, then add predictor which would produce the most benefit (biggest increase in $R^2$) when added to the existing model.
    b. If the p-value is large (over 5%), stop and discard this predictor. At this point, no (unused) predictor should be significant when added to the model and we are done.
    
- Why? Sometimes variables that were significant early on, can become insignificant after other new variables are added to the model.

Backward, stepwise selection is the same, except you perform forward selection every time you delete a term from the model.

## CAUTION {.smaller}

-   These automated methods have fallen out of favor in recent years, but you can still use them and should know what they are.
-   Automated methods ARE NOT a replacement for subject matter expertise
-   Think of the models that come out of these procedures as *suggestions*
-   The order in which variables are added to a model can *help* us understand which variables are more important and which are redundant.

Complete Exercise 5.

## Recap

-   Comparing subsets of models using nested F-Test
-   Choosing models using:
    -   Exhaustive search
    -   Forward/Backward/Stepwise selection