---
title: "Inference for Logistic Regression Models"
author: "Prof. Eric Friedlander"
date: "2024-11-13"
date-format: "MMM DD, YYYY"
footer: "[🔗 MAT 212 - Fall 2024 -  Schedule](https://mat212fa24.netlify.app/schedule)"
logo: "../images/logo.png"
format: 
  revealjs:
    theme: slides.scss
    multiplex: false
    transition: fade
    slide-number: false
    incremental: false 
    chalkboard: true
html-math-method:
  method: mathjax
  url: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js"
execute:
  freeze: auto
  echo: true
  cache: false
knitr:
  opts_chunk: 
    R.options:      
    width: 200
bibliography: references.bib
---

```{r}
#| include: false

# figure options
knitr::opts_chunk$set(
  fig.width = 8, fig.asp = 0.618, out.width = "90%",
  fig.retina = 3, dpi = 300, fig.align = "center"
)

library(countdown)
```

## Announcements {.midi}

-   Project: [Paper Due November 18th](https://mat212fa24.netlify.app/project-instructions#draft-report)
-   Oral R Quiz

::: appex
📋 [AE 21 - Inference for Logistic Regression Models](https://mat212fa24.netlify.app/ae/ae-21-logistic-inference)

- Open up AE 21 and complete Exercise 0.
:::



## Topics

::: nonincremental
-   Inference for coefficients in logistic regression
:::


## Computational setup

```{r}
#| echo: true

# load packages
library(tidyverse)
library(broom)
library(ggformula)
library(openintro)
library(knitr)
library(kableExtra)  # for table embellishments
library(Stat2Data)   # for empirical logit
library(countdown)

# set default theme and larger font size for ggplot2
ggplot2::theme_set(ggplot2::theme_bw(base_size = 20))
```

# Data

## Risk of coronary heart disease {.midi}

This data set is from an ongoing cardiovascular study on residents of the town of Framingham, Massachusetts. We want to examine the relationship between various health characteristics and the risk of having heart disease.

-   `TenYearCHD`:

    -   1: Developed heart disease in next 10 years
    -   0: Did not develop heart disease in next 10 years

-   `age`: Age at exam time (in years)

-   `currentSmoker`: 0 = nonsmoker, 1 = smoker

## Data prep

```{r}
heart_disease <- read_csv("data/framingham.csv") |>
  select(TenYearCHD, age, currentSmoker) |>
  drop_na() |>
  mutate(currentSmoker = as.factor(currentSmoker))

heart_disease |> head() |> kable()
```

## Modeling risk of coronary heart disease

Using `age`:

```{r}
risk_fit <- glm(TenYearCHD ~ age, 
      data = heart_disease, family = "binomial")

risk_fit |> tidy() |> kable()
```
    
# Inference for Logistic Regression

## Recall: Inference for Linear Regression

-   **t-test:** determine whether the a given $\beta_i$ (slope) was different that zero
-   **ANOVA/F-Test:** To test the full model or to compare nested models
-   **SSModel/SSE/ $R^2$ / $\hat{\sigma}_{\epsilon}$ :** metrics to try a measur the amount of variability explained by competing models

## Hypothesis test for $\beta_1$

**Hypotheses:** $H_0: \beta_1 = 0 \hspace{2mm} \text{ vs } \hspace{2mm} H_a: \beta_1 \neq 0$

. . .

**Test Statistic:** $$z = \frac{\hat{\beta}_1 - 0}{SE_{\hat{\beta}_1}}$$

$z$ is sometimes called a *Wald statistic* and this test is sometimes called a *Wald Hypothesis Test*.

. . .

**P-value:** $P(|Z| > |z|)$, where $Z \sim N(0, 1)$, the Standard Normal distribution

## Confidence interval for $\beta_1$

We can calculate the **C% confidence interval** for $\beta_1$ as the following:

$$
\Large{\hat{\beta}_1 \pm z^* SE_{\hat{\beta}_1}}
$$

where $z^*$ is calculated from the $N(0,1)$ distribution

. . .

::: callout-note
This is an interval for the change in the *log-odds* for every one unit increase in $x$
:::

## Interpretation in terms of the odds

The change in **odds** for every one unit increase in $x_1$.

$$
\Large{\exp\{\hat{\beta}_1 \pm z^* SE_{\hat{\beta}_1}\}}
$$

. . .

**Interpretation:** We are $C\%$ confident that for every one unit increase in $x_1$, the odds multiply by a factor of $\exp\{\hat{\beta}_1 - z^* SE_{\hat{\beta}_1}\}$ to $\exp\{\hat{\beta}_1 + z^* SE_{\hat{\beta}_1}\}$, holding all else constant.

## Coefficient for `age` {.midi}

```{r}
#| label: risk-model-age-highlight
#| echo: false

tidy(risk_fit, conf.int = TRUE) |> 
  kable(digits = 3) |>
  row_spec(2, background = "#D9E3E4")
```

. . .

**Hypotheses:**

$$
H_0: \beta_{age} = 0 \hspace{2mm} \text{ vs } \hspace{2mm} H_a: \beta_{age} \neq 0
$$

## Coefficient for `age` {.midi}

```{r}
#| echo: false
#| ref.label: risk-model-age-highlight
```

**Test statistic:**

$$z = \frac{0.0747 - 0}{0.00527} \approx 14.178$$

Note: rounding errors!

## Coefficient for `age` {.midi}

```{r}
#| echo: false
#| ref.label: risk-model-age-highlight
```

**P-value:**

$$
P(|Z| > |14.178|) \approx 0
$$

. . .

```{r}
2 * pnorm(14.178,lower.tail = FALSE)
```

## Coefficient for `age` {.midi}

```{r}
#| echo: false
#| ref.label: risk-model-age-highlight
```

**Conclusion:**

The p-value is very small, so we reject $H_0$. The data provide sufficient evidence that age is a statistically significant predictor of whether someone will develop heart disease in the next 10 years.

## CI for `age`

```{r}
#| echo: false
#| ref.label: risk-model-age-highlight
```

We are 95% confident that for each additional year of age, the change in the *log-odds* of someone developing heart disease in the next 10 years is between 0.064 and 0.085.

. . .

We are 95% confident that for each additional year of age, the *odds* of someone developing heart disease in the next 10 years will increase by a factor of $\exp(0.064) \approx 1.077$ to $\exp(0.085)\approx 1.089$.

. . .

Complete Exercises 1-4.


# Recap

## Inference: Linear vs. Logistic Regression

-   **t-test:** determine whether the a given $\beta_i$ (slope) was different that zero
-   **ANOVA/F-Test:** To test the full model or to compare nested models
-   **SSModel/SSE/ $R^2$ / $\hat{sigma}_{\epsilon}:** metrics to try a measur the amount of variability explained by competing models

## Inference: Linear vs. Logistic Regression

-   ~~**t-test:** determine whether the a given $\beta_i$ (slope) was different that zero~~
-   **ANOVA/F-Test:** To test the full model or to compare nested models (Next time)
-   **SSModel/SSE/ $R^2$ / $\hat{\sigma}_{\epsilon}$ :** metrics to try a measur the amount of variability explained by competing models (Next time)

## Inference for $\beta_1$

-   Same idea as linear regression
-   Define null and alternative hypotheses
    -   Null: my variable has no effect
    -   Alternative: my variable has an effect
-   Compute test statistic (how many standard errors is my observed $\hat{\beta}_1$ away from 0?)
-   Compute p-value
    -   What is the probability I got my $\hat{\beta}_1$ *or even stronger evidence of an effect* if the null hypothesis were the truth?



