---
title: "Inference for Logistic Regression Models Continued"
author: "Prof. Eric Friedlander"
date: "2024-11-15"
date-format: "MMM DD, YYYY"
footer: "[🔗 MAT 212 - Fall 2024 -  Schedule](https://mat212fa24.netlify.app/schedule)"
logo: "../images/logo.png"
format: 
  revealjs:
    theme: slides.scss
    multiplex: false
    transition: fade
    slide-number: false
    incremental: false 
    chalkboard: true
html-math-method:
  method: mathjax
  url: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js"
execute:
  freeze: auto
  echo: true
  cache: false
knitr:
  opts_chunk: 
    R.options:      
    width: 200
bibliography: references.bib
---

```{r}
#| include: false

# figure options
knitr::opts_chunk$set(
  fig.width = 8, fig.asp = 0.618, out.width = "90%",
  fig.retina = 3, dpi = 300, fig.align = "center"
)

library(countdown)
```

## Do now

Write your confidence interval interpretations from Wednesday on the board.

## Announcements {.midi}

-   Project: [Paper Due November 18th](https://mat212fa24.netlify.app/project-instructions#draft-report)
-   Oral R Quiz

::: appex
📋 [AE 22 - Inference for Logistic Regression Models Continued](https://mat212fa24.netlify.app/ae/ae-22-logistic-inference2)

- Open up AE 22 and complete Exercise 0.
:::



## Topics

::: nonincremental
-   Inference for coefficients in logistic regression
:::


## Computational setup

```{r}
#| echo: true

# load packages
library(tidyverse)
library(broom)
library(ggformula)
library(knitr)
library(kableExtra)

# set default theme and larger font size for ggplot2
ggplot2::theme_set(ggplot2::theme_bw(base_size = 20))
```

# Data

## Risk of coronary heart disease {.midi}

This data set is from an ongoing cardiovascular study on residents of the town of Framingham, Massachusetts. We want to examine the relationship between various health characteristics and the risk of having heart disease.

-   `TenYearCHD`:

    -   1: Developed heart disease in next 10 years
    -   0: Did not develop heart disease in next 10 years

-   `age`: Age at exam time (in years)


## Data prep

```{r}
heart_disease <- read_csv("data/framingham.csv") |>
  select(TenYearCHD, age) |>
  drop_na()

heart_disease |> head() |> kable()
```


## Modeling risk of coronary heart disease

Using `age`:

```{r}
risk_fit <- glm(TenYearCHD ~ age, data = heart_disease, 
                family = "binomial")

risk_fit |> tidy() |> kable()
```
    
# Inference for Logistic Regression

## Recall: Inference for Linear Regression

-   **t-test:** determine whether $\beta_1$ (the slope) is different than zero (**Last Time**)
-   **ANOVA/F-Test:** To test the full model or to compare nested models
-   **SSModel/SSE/ $R^2$ / $\hat{\sigma}_{\epsilon}$ :** metrics to try a measure the amount of variability explained by competing models

## Recall: Likelihood

$$
L = \prod_{i=1}^n\hat{\pi}_i^{y_i}(1 - \hat{\pi}_i)^{1 - y_i}
$$

-   Intuition: probability of obtaining our data given a certain set of parameters

$$
L(\hat{\beta}_0, \ldots, \hat{\beta}_p) = \prod_{i=1}^n\hat{\pi}_i(\hat{\beta}_0, \ldots, \hat{\beta}_p)^{y_i}(1 - \hat{\pi}_i(\hat{\beta}_0, \ldots, \hat{\beta}_p))^{1 - y_i}
$$

## Recall: Log-Likelihood

Taking the log makes the likelihood easier to work with and **doesn't change which $\beta$'s maximize it**.

$$
\log L = \sum\limits_{i=1}^n[y_i \log(\hat{\pi}_i) + (1 - y_i)\log(1 - \hat{\pi}_i)]
$$

## Log-Likelihood to Deviance

-   The log-likelihood measures of how well the model fits the data

-   Higher values of $\log L$ are better

-   **Deviance** = $-2 \log L$

    -   $-2 \log L$ follows a $\chi^2$ distribution with 1 degree of freedom
    
    -   Think of deviace as the analog of the residual sum of squares (SSE) in linear regression



## Calculate deviance for our model:

We can use our trusty ol' `glance` function

```{r}
risk_fit |>  glance() |> kable()
```

. . .

Complete Exercise 1.

## Comparing nested models

-   Suppose there are two models:

    -   Reduced Model includes only an intercept $\beta_0$
    -   Full Model includes $\beta_1$

-   We want to test the hypotheses

    $$
    \begin{aligned}
    H_0&: \beta_{1} = 0\\
    H_A&: \beta_1 \neq 0
    \end{aligned}
    $$

-   To do so, we will use something called a **Likelihood Ratio test (LRT)**, also known as the **Drop-in-deviance test**

## Likelihood Ratio Test (LRT) {.smaller}

**Hypotheses:**

$$
\begin{aligned}
H_0&: \beta_1 = 0 \\
H_A&: \beta_1 \neq 0
\end{aligned}
$$

. . .

**Test Statistic:** $$G = (-2 \log L_{reduced}) - (-2 \log L_{full})$$

Sometimes written as $$G = (-2 \log L_0) - (-2 \log L)$$

. . .

**P-value:** $P(\chi^2 > G)$, calculated using a $\chi^2$ distribution with 1 degree of freedom

## $\chi^2$ distribution

```{r}
#| echo: false
#| fig-height: 6

x <- seq(from =0, to = 10, length = 100)

# Evaluate the densities
y_1 <- dchisq(x, 1)
y_2 <- dchisq(x,2)
y_3 <- dchisq(x,3)
y_4 <- dchisq(x,5)

# Plot the densities
plot(x, y_1, col = 1, type = "l", ylab="",lwd=3, ylim = c(0, 0.5), 
     main  = "Chi-square Distribution")
lines(x,y_2, col = 2,lwd=3)
lines(x, y_3, col = 3,lwd=3)
lines(x, y_4, col = 4,lwd=3)

# Add the legend
legend("topright",
       c("df = 1", "df = 2 ", "df = 3", "df = 5"), 
       col = c(1, 2, 3, 4), lty = 1)
```

## Reduced model {.smaller}

First model, reduced:

```{r}

risk_fit_reduced <- glm(TenYearCHD ~ 1, 
      data = heart_disease, family = "binomial",
      control = glm.control(epsilon = 1e-20)) # Ignore this line

risk_fit_reduced |> tidy() |> kable()
```

. . .

Side bar... 

::::{.columns}
:::{.column}
Probability predicted by model

```{r}
exp(coef(risk_fit_reduced)[1])/(1 + exp(coef(risk_fit_reduced)[1]))
```
:::
:::{.column}
```{r}
library(mosaic)
tally(~TenYearCHD, data = heart_disease, format = "proportion") |> kable()
```
:::
::::

## Should we add `age` to the model? {.smaller}

Second model, full:

```{r}
risk_fit_full <- glm(TenYearCHD ~ age, 
      data = heart_disease, family = "binomial")

risk_fit_full |>  tidy() |> kable()
```

. . .

Calculate deviance for each model:

```{r}
dev_reduced <- glance(risk_fit_reduced)$deviance #Use $ instead of select

dev_full <- glance(risk_fit_full)$deviance

dev_reduced

dev_full
```

. . .

Drop-in-deviance test statistic:

```{r}
test_stat <- dev_reduced - dev_full
```

## Should we add `age` to the model?

Calculate the p-value using a `pchisq()`, with 1 degree of freedom:

```{r}
pchisq(test_stat, 1, lower.tail = FALSE)
```

. . .

**Conclusion:** The p-value is very small, so we reject $H_0$. The data provide sufficient evidence that the coefficient of `age` is not equal to 0. Therefore, we should add it to the model.

. . .

Complete Exercises 2-3.

## Drop-in-Deviance test in R {.midi}

-   We can use the **`anova`** function to conduct this test

-   Add **`test = "Chisq"`** to conduct the drop-in-deviance test

. . .

```{r}
anova(risk_fit_reduced, risk_fit_full, test = "Chisq") |>
  tidy() |> kable(digits = 3)
```

Complete Exercises 4-5, 6 (time permitting)



# Recap

## Inference: Linear vs. Logistic Regression

-   [**t-test:** determine whether $\beta_1$ (the slope) is different than zero]{.fragment .strike}
-   [**ANOVA/F-Test:** To test the full model or to compare nested models]{.fragment .semi-fade-out}
-   [**SSModel/SSE/ $R^2$ / $\hat{\sigma}_{\epsilon}$ :** metrics to try and measure the amount of variability explained by competing models]{.fragment .semi-fade-out}

## Inference for $\beta_1$

-   Last time: Wald Test
-   Likelihood Ratio Test
    -   More reliable than Wald
    -   More computationally taxing
    -   Deviance: think of like SSE
-   Next time: Multiple predictors!


