---
title: "Multiple Logistic Regression"
subtitle: "Selection + Transformations"
author: "Prof. Eric Friedlander"
date: "2024-12-02"
date-format: "MMM DD, YYYY"
footer: "[🔗 MAT 212 - Fall 2024 -  Schedule](https://mat212fa24.netlify.app/schedule)"
logo: "../images/logo.png"
format: 
  revealjs:
    theme: slides.scss
    multiplex: false
    transition: fade
    slide-number: false
    incremental: false 
    chalkboard: true
html-math-method:
  method: mathjax
  url: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js"
execute:
  freeze: auto
  echo: true
  cache: false
knitr:
  opts_chunk: 
    R.options:      
    width: 200
bibliography: references.bib
---

```{r}
#| include: false

# figure options
knitr::opts_chunk$set(
  fig.width = 8, fig.asp = 0.618, out.width = "90%",
  fig.retina = 3, dpi = 300, fig.align = "center"
)

library(countdown)
```


## Announcements

-   [Exam 2 Corrections due Friday in class](https://cofi.instructure.com/courses/15692/assignments/201717)

-   [Final Paper due Monday, December 9th](https://mat212fa24.netlify.app/project-instructions#written-report)

-   [Virtual Poster due Tuesday, December 10th 11:59pm](https://mat212fa24.netlify.app/project-instructions#poster-presentation)

-   [Poster Presentation Wednesday, December 11th 8:30-11am CML 105](https://mat212fa24.netlify.app/project-instructions#poster-presentation)

::: appex
📋 [AE 24 - Logistic Regression: Selection and Comparison](https://mat212fa24.netlify.app/ae/ae-24-comparing-logistic)

- Open up AE 24.
:::

## Topics

-   Investigating interaction terms in logistic regression models

-   Comparing logistic regression models

-   Choosing logistic regression models

-   Transformations in logistic regression

## Computational setup

```{r}
#| echo: true

# load packages
library(tidyverse)
library(broom)
library(openintro)
library(knitr)
library(kableExtra)  # for table embellishments
library(Stat2Data)   # for empirical logit

# set default theme and larger font size for ggplot2
ggplot2::theme_set(ggplot2::theme_bw(base_size = 20))
```

# Data

## Risk of coronary heart disease {.midi}

This data set is from an ongoing cardiovascular study on residents of the town of Framingham, Massachusetts. We want to examine the relationship between various health characteristics and the risk of having heart disease.

-   `TenYearCHD`:

    -   1: Developed heart disease in next 10 years
    -   0: Did not develop heart disease in next 10 years

-   `age`: Age at exam time (in years)

-   `education`: 1 = Some High School, 2 = High School or GED, 3 = Some College or Vocational School, 4 = College

## Data prep

```{r}
heart_disease <- read_csv("data/framingham.csv") |>
  select(TenYearCHD, age, education, male, diaBP) |>
  drop_na() |> 
  mutate(education = as.factor(education),
         male = as.factor(male))
         

heart_disease |> head() |> kable()
```
# Interaction Terms

## Interactions

**Intuitively:** There is an interaction between two variables when the influence of one variables on the log-odds depends on what the value of the other variable is.

. . .

Complete Exercise 1. 

```{r}
countdown::countdown(minutes = 1, seconds=30)
```
## Recall: Empirical Logit Plot

```{r}
emplogitplot1(TenYearCHD ~ age, data = heart_disease,
              ngroups = 10)
```

:::{.question}
How do we interpret this?
:::


## Visualizing Interactions (Quant vs. Cat)

```{r}
emplogitplot2(TenYearCHD ~ age + male, data = heart_disease,
              ngroups = 10)
```

. . .

Complete Exercise 2.

## Visualizing Interactions (Quant vs. Cat)

```{r}
emplogitplot2(TenYearCHD ~ age + education, data = heart_disease,
              ngroups = 10)
```

:::{.question}
Based on this plot, do you think there is an interaction between age and education?
:::

## Models with Interaction 

```{r}
#| code-fold: true
age_male_interaction_model <- glm(TenYearCHD ~ age + male + age*male, 
                  data = heart_disease, family = "binomial")

age_male_interaction_model |> tidy() |> kable(digits = 3)
```

## Models with Interaction {.smaller}

Two model's in one!

-   Non-male's: $$\log\left(\frac{\hat{p}}{1-\hat{p}}\right) = -6.358 +0.085\times\text{age}$$
-   Male's: 
$$
\begin{aligned} \log\left(\frac{\hat{p}}{1-\hat{p}}\right) &= (-6.358+1.308) +(0.085-0.015)\times\text{age}\\
&=  -5.05 +0.070\times\text{age}
\end{aligned}
$$

. . .

Complete Exercises 3 and 4.


# Model selection

## AIC or BIC for model selection

$$
\begin{align}
&AIC = - 2 * \log L - {\color{purple}n\log(n)}+ 2(p +1)\\[5pt]
&BIC =- 2 * \log L - {\color{purple}n\log(n)} + \log(n)\times(p+1)
\end{align}
$$

## AIC from the `glance()` function

Let's look at the AIC for model we fit earlier

```{r}
glance(age_male_interaction_model)$AIC
```

. . .

**Calculating AIC**

```{r}
- 2 * glance(age_male_interaction_model)$logLik + 2 * (3+1)
```

## Comparing the models using AIC

Let's compare a few models:

```{r}
age_model <- glm(TenYearCHD ~ age, data = heart_disease, family = "binomial")
age_male_model <- glm(TenYearCHD ~ age+male, data = heart_disease, family = "binomial") 
age_education_model <-  glm(TenYearCHD ~ age+education, data = heart_disease, family = "binomial")
age_male_education_model <-  glm(TenYearCHD ~ age+ male + education, data = heart_disease, family = "binomial")
age_education_interaction_model <-  glm(TenYearCHD ~ age + education + age* education, data = heart_disease, family = "binomial")
all_interaction_model <-  glm(TenYearCHD ~ (age + education + male)^2, data = heart_disease, family = "binomial")
```

## Comparing the models using AIC {.smaller}

Let's compare a few models:

```{r}
aics <- c(glance(age_model)$AIC,
glance(age_male_model)$AIC,
glance(age_male_interaction_model)$AIC,
glance(age_education_model)$AIC,
glance(age_male_education_model)$AIC,
glance(age_education_interaction_model)$AIC,
glance(all_interaction_model)$AIC)

aics
```
<br>

::: question
Based on AIC, which model would you choose? Raise you hand if your AIC is better.
:::

. . .

```{r}
which.min(aics)
```



## Comparing the models using BIC {.smaller}

Let's compare the models using BIC

```{r}
bics <- c(glance(age_model)$BIC,
glance(age_male_model)$BIC,
glance(age_male_interaction_model)$BIC,
glance(age_education_model)$BIC,
glance(age_male_education_model)$BIC,
glance(age_education_interaction_model)$BIC,
glance(all_interaction_model)$BIC)

bics
```

<br>

::: question
Based on BIC, which model would you choose? Raise your hand if your BIC is better.

:::

. . .

```{r}
which.min(bics)
```

## Choosing a Logistic Regression Model

-   Forward/Backward/Stepwise selection all work in the same way
-   Can string together a sequence of Likelihood Ratio Tests (but only a few)

# Transformations

## Power Family of Transformation

-   Many departures from linearity can be solved with power transformations (e.g. $X^{power}$)

    +   For technical reasons, $power = 0$ corresponds to $\log$
    
-   Look at empirical log-odds plot

-   Concave down pattern $\Rightarrow$ transform down (i.e. $power < 1$)

    +   $\log$ is typically a good first choice

-   Concave up pattern $\Rightarrow$ transform up (i.e. $power > 1$)


## Empirical Log-Odds Plot

```{r}
emplogitplot1(TenYearCHD ~ diaBP, data = heart_disease, ngroups = 4)
```

## Empirical Log-Odds Plot

```{r}
emplogitplot1(TenYearCHD ~ diaBP, data = heart_disease, ngroups = 4)
```

:::{.question}
Concave up or down?
:::

## Empirical Log-Odds Plot

```{r}
heart_disease <- heart_disease |> mutate(diaBP_transformed = diaBP^2)
emplogitplot1(TenYearCHD ~ diaBP_transformed, data = heart_disease, ngroups = 4)
```

## Empirical Log-Odds Plot

```{r}
heart_disease <- heart_disease |> mutate(diaBP_transformed = diaBP^10)
emplogitplot1(TenYearCHD ~ diaBP_transformed, data = heart_disease, ngroups = 4)
```

## Empirical Log-Odds Plot

```{r}
heart_disease <- heart_disease |> mutate(diaBP_transformed = diaBP^6)
emplogitplot1(TenYearCHD ~ diaBP_transformed, data = heart_disease, ngroups = 4)
```




## Recap

-   Interaction terms in logistic regression models

    +   For quantitative-categorical, creates nested models

-   Comparing logistic regression models

    +   AIC/BIC

-   Choosing logistic regression models

    +   Same selection techniques as linear regression

-   Transformations in logistic regression

    +   Can only transform $X$
  
    +   Concave [up/down] $\Rightarrow$ power-transform [up/down]

