---
title: "AE 12: Categorical Predictors"
subtitle: "Pee-to-Peer Loans"
editor: visual
format: 
  pdf:
    pdf-engine: pdflatex
---

::: callout-important
-   Open [RStudio](https://rstudio.collegeofidaho.edu/) and create a subfolder in your AE folder called "AE-12".

-   Go to the [Canvas](https://cofi.instructure.com/courses/15692/assignments/183202) and locate your `AE-12` assignment to get started.

-   Upload the `ae-12.qmd` file into the folder you just created. The `.qmd` and PDF responses are due in Canvas. You can check the due date on the Canvas assignment.
:::

## Packages + data

```{r load-packages}
#| message: false
library(tidyverse)
library(ggformula)
library(mosaic)
library(broom)
library(knitr)
library(openintro)

loan50 <- loan50 |>
  mutate(annual_income_k = annual_income / 1000)
```

The data for this AE is is a sample of 10,000 loans made through a peer-to-peer lending club. The data is in the `loan50` data frame in the **openintro** R package.

## Variables

-   `annual_income_k`: Annual income in \$1,000's
-   `verified_income`: Whether borrower's income source and amount have been verified (`Not Verified`, `Source Verified`, `Verified`)

**Response**: `interest_rate`: Interest rate for the loan

## Analysis goal

-   Predict `interest_rate` using the categorical variable `verified_income` as a predictor
-   Include other quantitative variables and understand how they interact with `verified_income`

## Exercise 1

Generate side-by-side boxplots of `interest_rate` vs. `verified_income`. Does it appear that there is a relationship between the two variables?

## Exercise 2

Based on the output of the code below, what do you think would be the best predictions for the interest rate of a borrow with `Not Verified`, `Source Verified`, and `Verified` income, respectively.

```{r}
favstats(interest_rate ~ verified_income, data = loan50) |> 
  kable()
```

## Exercise 3

Fit a linear model predicting `interest_rate` from `verified_income`. What is the reference level for `verified_income`?

## Exercise 4

WITHOUT WRITING ANY CODE except for addition, subtraction, multiplication, and addition, what would the model predict the average `interest_rate` for each of the three levels of `verified_income`? How do these answers compare to your answers from Exercise 2?

## Exercise 5

The linear model below predicts `interest_rate` from `annual_income_k`. Add `verified_income` as a predictor to this model. Do not include an interaction term. Be prepared to discuss how and why the plot changes when you add in `verified_income`.

```{r}
ex5_model <- lm(interest_rate ~ annual_income_k, data = loan50) 
ex5_model |> 
  tidy() |> 
  kable()

plotModel(ex5_model) # nifty function from the mosaic package
```

## Exercise 6

How do you think the plot above will change if you add in an interaction term between `verified_income` and `interest_rate`? **AFTER** thinking about it, add in an interaction term between `verified_income` and `annual_income_k`.

```{r}
ex6_model <- lm(interest_rate ~ annual_income_k, data = loan50) 
ex6_model |> 
  tidy() |> 
  kable()

plotModel(ex6_model) # nifty function from the mosaic package
```

## Exercise 7

Based on the model above (and the equation on the slides):

-   Write the equation of the model to predict interest rate for applicants with *Not Verified* income.
-   Write the equation of the model to predict interest rate for applicants with *Verified* income.
-   Our degrees of freedom will be $n-p-1$. What is $p$ in this case? Hint: it isn't 2.

## To submit the AE

::: callout-important
-   Render the document to produce the PDF with all of your work from today's class.
-   Upload your QMD and PDF files to the Canvas assignment.
:::
