---
title: "AE 09: Outliers"
subtitle: "Baseball game times"
author: "Driver: _______, Reporter: _______, Gopher: ________"
editor: visual
format: 
  pdf:
    pdf-engine: pdflatex
---

::: callout-important
-   Open [RStudio](https://rstudio.collegeofidaho.edu/) and create a subfolder in your AE folder called "AE-09".

-   Go to the [Canvas](https://cofi.instructure.com/courses/15692/assignments/182299) and locate your `AE-09` assignment to get started.

-   Upload the `ae-09.qmd` file into the folder you just created. The `.qmd` and PDF responses are due in Canvas. You can check the due date on the Canvas assignment.
:::

```{r}
#| label: load-packages
#| warning: false
#| message: false
library(tidyverse)
library(ggformula)
library(yardstick)
library(Stat2Data)
library(mosaic)
library(broom)
library(knitr)
library(patchwork) #arrange plots in a grid
```

## Data

The data set for this assignment is from the `Stat2Data` R package which is the companion package for this course's textbook. The data contains data from all MLB games played on August, 11, 2017. On this day there were no extra-innings games or rain delays. You can find information [here](https://cran.r-project.org/web/packages/Stat2Data/Stat2Data.pdf) by searching for the *Baseball Game TImes of One Day in 2017* dataset.

```{r}
#| label: load-data
#| warning: false
#| message: false

data("BaseballTimes2017") # Loads the data from the package
```

We are interested in predicting `Time` the time in minutes to play the game, from either `Runs`, the number of runs scores by the two teams combined, or `Pitchers`, the number of pitchers sued total for the two teams.

## Exercise 1

Argue that `Runs` is a better predictor of `Time` than `Pitchers`.

## Exercise 2

Argue whether you think the CIN-MIL game would be considered a high leverage and/or high influence point.

## Exercise 3

Remove the CIN-MIL game from the data set. Which model is better now?

## To submit the AE:

::: callout-important
-   Render the document to produce the PDF with all of your work from today's class.
-   Upload your QMD and PDF files to the Canvas assignment.
:::
