---
title: "AE 01: Movie Budgets and Revenues"
date: "August 28, 2024"
editor: visual
---

::: callout-important
For this AE, you will discuss the questions in groups and submit answers on Canvas. This AE does not count towards the Application Exercise grade.
:::

We will look at the relationship between budget and revenue for movies made in the United States in 1986 to 2020. The dataset is created based on data from the [Internet Movie Database (IMDB)](https://www.imdb.com/).

```{r include = FALSE}
knitr::opts_chunk$set(warning = FALSE, 
                      message = FALSE)
```

```{r load-packages}
#| warning: FALSE
#| message: FALSE
# library(readr)
library(tidyverse)
library(ggformula)
```

## Data

The `movies` data set includes basic information about each movie including budget, genre, movie studio, director, etc. A full list of the variables may be found [here](https://github.com/danielgrijalva/movie-stats).

```{r read-data, message=FALSE, warning=FALSE}
movies <- read_csv("https://raw.githubusercontent.com/danielgrijalva/movie-stats/master/movies.csv")
```

View the first 10 rows of data.

```{r}
movies |>
  slice(1:10)
```

## Analysis

We begin by looking at how the average gross revenue (`gross`) has changed over time. Since we want to visualize the results, we will choose a few genres of interest for the analysis.

```{r genre-list}
genre_list <- c("Comedy", "Action", "Animation", "Horror")
```

```{r plot-rating}
#| warning: FALSE
#| fig-width: 10

movies |>
  filter(genre %in% genre_list) |> 
  group_by(genre,year) |>
  summarise(avg_gross = mean(gross)) |>
  gf_point(avg_gross ~ year, color = ~genre) |>
  gf_line() |> 
  gf_labs(y = "Average Gross Revenue (in US Dollars)",
    title = "Gross Revenue Over Time")
```

::: {.callout-note appearance="simple"}
*What do you observe from the plot?*
:::

Next, let's see the relationship between a movie's budget and its gross revenue.

```{r plot-revenue, fig.width=10, warning=FALSE}
movies |>
  filter(genre %in% genre_list, budget > 0) |> 
  gf_point(log(gross) ~ log(budget), color = ~genre) |> 
  gf_lm(se=FALSE) |>  
  gf_labs(x = "Log-transformed Budget",
  y = "Log-transformed Gross Revenue") |> 
  gf_facet_wrap(~ genre)
```

## Exercises

Please enter your names and responses in [Canvas](https://cofi.instructure.com/courses/15692/discussion_topics/65535).

1.  Suppose we fit a regression model for each genre that uses budget to predict gross revenue. What are the signs of the correlation between `budget` and `gross` and the slope in each regression equation?

2.  Suppose we fit the regression model from the previous question. Which genre would you expect to have the smallest residuals, on average (*residual = observed revenue - predicted revenue*)?

3.  \[Time permitting\] Discuss the following: Notice in the graph above that `budget` and `gross` are log-transformed. Why are the log-transformed values of the variables displayed rather than the original values (in U.S. dollars)?

## References

1.  [github.com/danielgrijalva/movie-stats](https://github.com/danielgrijalva/movie-stats){.uri}
2.  [Internet Movie Database](https://www.imdb.com/)

## Appendix

Below is a list of genres in the data set:

```{r}
movies |> 
  arrange(genre) |> 
  select(genre) |>
  distinct()
```
