---
title: "AE 07: Model evaluation"
subtitle: "Songs on Spotify"
author: "Driver: _______, Reporter: _______, Gopher: ________"
editor: visual
---

::: callout-important
-   Open [RStudio](https://rstudio.collegeofidaho.edu/) and create a subfolder in your AE folder called "AE-07".

-   Go to the [Canvas](https://cofi.instructure.com/courses/15692/assignments/182123) and locate your `AE 07` assignment to get started.

-   Upload the `ae-07.qmd` and `spotify-popular.csv` files into the folder you just created. The `.qmd` and PDF responses are due in Canvas. You can check the due date on the Canvas assignment.
:::

```{r}
#| label: load-packages
#| warning: false
#| message: false
library(tidyverse)
library(ggformula)
library(mosaic)
library(broom)
library(knitr)
library(patchwork) #arrange plots in a grid
```

## Data

The data set for this assignment is a subset from the [Spotify Songs](https://github.com/rfordatascience/tidytuesday/blob/master/data/2020/2020-01-21/readme.md) Tidy Tuesday data set. The data were originally obtained from Spotify using the **spotifyr** R package.

It contains numerous characteristics for each song. You can see the full list of variables and definitions [here](https://github.com/rfordatascience/tidytuesday/blob/master/data/2020/2020-01-21/readme.md). This analysis will focus specifically on the following variables:

| variable         | class     | description                                                                                                                                                                                                                                                                                                                                                                               |
|:--------|:--------|:------------------------------------------------------|
| track_id         | character | Song unique ID                                                                                                                                                                                                                                                                                                                                                                            |
| track_name       | character | Song Name                                                                                                                                                                                                                                                                                                                                                                                 |
| track_artist     | character | Song Artist                                                                                                                                                                                                                                                                                                                                                                               |
| track_popularity | double    | Song Popularity (0-100) where higher is better                                                                                                                                                                                                                                                                                                                                            |
| energy           | double    | Energy is a measure from 0.0 to 1.0 and represents a perceptual measure of intensity and activity. Typically, energetic tracks feel fast, loud, and noisy. For example, death metal has high energy, while a Bach prelude scores low on the scale. Perceptual features contributing to this attribute include dynamic range, perceived loudness, timbre, onset rate, and general entropy. |
| valence          | double    | A measure from 0.0 to 1.0 describing the musical positiveness conveyed by a track. Tracks with high valence sound more positive (e.g. happy, cheerful, euphoric), while tracks with low valence sound more negative (e.g. sad, depressed, angry).                                                                                                                                         |

```{r}
#| label: load-data
#| warning: false
#| message: false
spotify <- read_csv("data/spotify-popular.csv")
```

Are high energy songs more positive? To answer this question, we'll analyze data on some of the most popular songs on Spotify, i.e. those with `track_popularity >= 80`. We'll use linear regression to fit a model to predict a song's positiveness (`valence`) based on its energy level (`energy`).

Below are plots as part of the exploratory data analysis.

```{r}
#| label: univariate-eda
#| warning: false
p1 <- gf_histogram(~valence, data = spotify) |>  
  gf_labs(title = "Distribution of Valence", 
       subtitle = " for Popular songs on Spotify", 
       x = "Valence")

p2 <- gf_histogram(~energy, data = spotify) |> 
  gf_labs(title = "Distribution of Energy", 
       subtitle = "for Popular songs on Spotify", 
       x = "Energy")
p1 + p2
```

```{r}
#| label: bivariate-eda
#| warning: false
gf_point(valence ~ energy, data = spotify) |> 
  gf_labs(title = "Valence vs. Energy", 
       subtitle = "Popular songs on Spotify", 
       x = "Energy", 
       y = "Valence")
```

## Model with 90% CI for coefficients

```{r}
#| label: fit-model
#| echo: true

spotify_fit <- lm(valence ~ energy, data = spotify)

tidy(spotify_fit, conf.int = TRUE, conf.level = 0.9) |> 
  kable(digits = 3)
```

## Create augmented data frame

```{r}
#| label: augment-data
#| echo: true

spotify_aug <- augment(spotify_fit)
```

## Create residual plots

```{r}
#| label: resid-vs-fitted
#| echo: true
#| warning: false

gf_point(.resid ~ .fitted, data = spotify_aug) |>  
  gf_hline(yintercept = 0, color = "red", linetype = "dashed") |> 
  gf_labs(x = "Fitted values", 
       y = "Residuals", 
       title = "Residuals vs. fitted")
```

```{r}
#| label: resid-dist
#| warning: false

resid_hist <- gf_histogram(~.resid, data = spotify_aug) |>  
  gf_labs(x = "Residuals", 
       title = "Distribution of residuals")

resid_qq <- gf_qq(~.resid, data = spotify_aug) |> 
  gf_qqline()  |> 
  gf_labs(x = "Theoretical quantiles",
       y = "Observed residuals", 
       title = "Normal QQ-plot")

resid_hist + resid_qq
```

## Exercise 1

Calculate $R^2$ and interpret this value in the context of the data.

```{r}
#| label: calc-rsq
```

## Exercise 2

Calculate RMSE and interpret this value in the context of the data.

```{r}
#| label: calc-rmse
```

## Exercise 3

Choose another quantitative variable in the data set. Check [here](https://github.com/rfordatascience/tidytuesday/blob/master/data/2020/2020-01-21/readme.md) for specific definitions of variables. After Dr. Friedlander approves your choice, fit a model predicting `valence` using that variable as a predictor. Make an argument, citing quantitative evidence, for which model is better, the energy model or your new model. Reporters, be ready to discuss with the class.

## To submit the AE:

::: callout-important
- Render the document to produce the PDF with all of your work from today's class.
- Upload your QMD and PDF files to the Canvas assignment.
:::
