---
title: "AE 05: Mathematical Models"
subtitle: "Songs on Spotify"
author: "Driver: _______, Reporter: _______, Gopher: ________"
date: "Sep 11, 2024"
editor: visual
execute: 
  warning: false
  message: false
---

::: callout-important
-   Open [RStudio](https://rstudio.collegeofidaho.edu/) and create a subfolder in your AE folder called "AE-05"

-   Go to the [Canvas](https://cofi.instructure.com/courses/15692/assignments/181903) and locate your `AE 05` assignment to get started.

-   Upload the `ae-05.qmd` and `spotify-popular.csv` files into the folder you just created. The `.qmd` and PDF responses are due in Canvas no later than Saturday, September 14 at 11:59pm.
:::

```{r}
#| label: load-packages
#| warning: false
#| message: false
library(tidyverse)
library(ggformula)
library(broom)
library(knitr)
library(patchwork) #arrange plots in a grid
```

## Data

The data set for this assignment is a subset from the [Spotify Songs](https://github.com/rfordatascience/tidytuesday/blob/master/data/2020/2020-01-21/readme.md) Tidy Tuesday data set. The data were originally obtained from Spotify using the **spotifyr** R package.

It contains numerous characteristics for each song. You can see the full list of variables and definitions [here](https://github.com/rfordatascience/tidytuesday/blob/master/data/2020/2020-01-21/readme.md). This analysis will focus specifically on the following variables:

| variable         | class     | description                                                                                                                                                                                                                                                                                                                                                                               |
|:-----------------|:-----------------|:-----------------------------------|
| track_id         | character | Song unique ID                                                                                                                                                                                                                                                                                                                                                                            |
| track_name       | character | Song Name                                                                                                                                                                                                                                                                                                                                                                                 |
| track_artist     | character | Song Artist                                                                                                                                                                                                                                                                                                                                                                               |
| track_popularity | double    | Song Popularity (0-100) where higher is better                                                                                                                                                                                                                                                                                                                                            |
| energy           | double    | Energy is a measure from 0.0 to 1.0 and represents a perceptual measure of intensity and activity. Typically, energetic tracks feel fast, loud, and noisy. For example, death metal has high energy, while a Bach prelude scores low on the scale. Perceptual features contributing to this attribute include dynamic range, perceived loudness, timbre, onset rate, and general entropy. |
| valence          | double    | A measure from 0.0 to 1.0 describing the musical positiveness conveyed by a track. Tracks with high valence sound more positive (e.g. happy, cheerful, euphoric), while tracks with low valence sound more negative (e.g. sad, depressed, angry).                                                                                                                                         |

```{r}
#| label: load-data
#| warning: false
#| message: false
spotify <- read_csv("data/spotify-popular.csv")
```

Are high energy songs more positive? To answer this question, we'll analyze data on some of the most popular songs on Spotify, i.e. those with `track_popularity >= 80`. We'll use linear regression to fit a model to predict a song's positiveness (`valence`) based on its energy level (`energy`).

Below are plots as part of the exploratory data analysis.

```{r}
#| label: univariate-eda
p1 <- gf_histogram(~valence, data = spotify) |>  
  gf_labs(title = "Distribution of Valence", 
       subtitle = " for Popular songs on Spotify", 
       x = "Valence")

p2 <- gf_histogram(~energy, data = spotify) |> 
  gf_labs(title = "Distribution of Energy", 
       subtitle = "for Popular songs on Spotify", 
       x = "Energy")
p1 + p2 # The patchwork package will arrange your plots for you
```

```{r}
#| label: bivariate-eda
gf_point(valence ~ energy, data = spotify) |> 
  gf_labs(title = "Valence vs. Energy", 
       subtitle = "Popular songs on Spotify", 
       x = "Energy", 
       y = "Valence")
```

## Exercise 1

Fit a model using the `energy` of a song to predict its `valence`, i.e. positiveness. Include the 90% confidence interval for the coefficients, and display the output using 3 digits.

```{r fit-model}
## add code
```

## Exercise 2

In words interpret the estimate and confidence interval for the slope in the previous exercise.

## Exercise 3

Interpret the p-value from Exercise 1.

## Exericse 4

Predict what the average valence for a song with an energy score 0.5 is. Report and interpret a 90% confidence interval for the average valence.

## Exericse 5

Report and interpret a 90% confidence interval for a single song with energy score 0.8.


::: callout-important
To submit the AE:

-   Render the document to produce the PDF with all of your work from today's class.
-   Upload your qmd and pdf files to the Canvas assignment.
:::
