---
title: "AE 20: Assessing Logistic Regression Models"
editor: visual
execute: 
  warning: false
  message: false
---

::: callout-important
-   Open [RStudio](https://rstudio.collegeofidaho.edu/) and create a subfolder in your AE folder called "AE-20".

-   Go to the [Canvas](https://cofi.instructure.com/courses/15692/assignments/184526) and locate your `AE-20` assignment to get started.

-   Upload the `ae-20.qmd` and `framingham.csv` files into the folder you just created. The `.qmd` and PDF responses are due in Canvas. You can check the due date on the Canvas assignment.
:::

## Packages

```{r}
#| label: load-pkgs-data
#| message: false
 
library(tidyverse)
library(broom)
library(ggformula)
library(mosaic)
library(knitr)

heart_disease <- read_csv("data/framingham.csv") |>
  select(totChol, TenYearCHD, age, BMI, cigsPerDay, heartRate, sysBP, diabetes) |>
  drop_na()
```

### Data: Framingham study

This data set is from an ongoing cardiovascular study on residents of the town of Framingham, Massachusetts. We want to predict if a randomly selected adult is high risk for heart disease in the next 10 years.

### Response variable

-   `TenYearCHD`:
    -   1: Patient developed heart disease within 10 years of exam
    -   0: Patient did not develop heart disease within 10 years of exam

### What's my predictor variable?

Based on your group, use the following as your predictor variable.

-   Group 1 - `totChol`: total cholesterol (mg/dL)
-   Group 2 -`BMI`: patient's body mass index
-   Group 3 -`cigsPerDay`: number of cigarettes patient smokes per day
-   Group 4 -`heartRate`: Heart rate (beats per minute)

### Additional Variables

-   `sysbp` - the patients systolic blood pressure at the time of examination

-   `diabetes` - 1 if the patient had diabetes and 0 if the patient didn't have diabetes at the time of examination

## Exercise 0

Fit a logistic regression model predicting `TenYearCHD` from your group's predictor variable. Note that we won't return to this model until Exercise 3.

## Exercise 1

Before you work on your own model, lets consider the variable `sysBP` which represents the patients systolic blood pressure (the top number). You have 120 seconds, find the $\beta$'s that result in the largest log-likelihood for a logistic regression model predicting the risk of coronary heart disease from `sysBP`:

```{r}
# Change these
beta0 <- 1
beta1 <- 1


# Don't change anything below this line
predicted_probabilities <- exp(beta0 + beta1*heart_disease$sysBP)/(1+exp(beta0 + beta1*heart_disease$sysBP))
log_likelihoods <- heart_disease$TenYearCHD*log(predicted_probabilities) +
                        (1-heart_disease$TenYearCHD)*log(1-predicted_probabilities)

# Final log likelihood
sum(log_likelihoods)

```

## Exercise 2

Compute the empirical logit for each level of `diabetes`:

1.  Use the function `tally` to compute the count the number of successes and failures for each level of `diabetes`.
2.  Compute the empirical odds.
3.  Compute the log of these odds.

## Exercise 3

Is linearity satisfied for the model you fit in Exercise 0?

## Exercise 4 (Time Permitting)

On a whiteboard, design a simulation-based hypothesis test for determining whether the coefficient of your explanatory variable is a statistically significant predictor for the risk of `TenYearCHD`.

## Submission

::: callout-important
To submit the AE:

-   Render the document to produce the PDF with all of your work from today's class.
-   Upload your QMD and PDF files to the Canvas assignment.
:::
