---
title: "Exam 02 Review"
date: "Nov 25, 2024"
editor: visual
execute: 
  warning: false
  message: false
---

::: callout-important
Imagine that this is printed out for you and all answers are given in written form.

Note: This in-class review is **not** exhaustive. Use lecture notes notes, application exercises, and homework for a comprehensive exam review. It's length is also not indicative of the length of the exam.
:::

## Packages

```{r}
#| message: false
 
library(tidyverse)
library(ggformula)
library(Stat2Data)
library(broom)
library(knitr)
library(rms)
```

## Data

The data for this analysis is about credit card customers. The following variables are in the data set:

-   `income`: Income in \$1,000's
-   `limit`: Credit limit
-   `rating`: Credit rating
-   `cards`: Number of credit cards
-   `age`: Age in years
-   `education`: Number of years of education
-   `own`: A factor with levels `No` and `Yes` indicating whether the individual owns their home
-   `student`: A factor with levels `No` and `Yes` indicating whether the individual was a student
-   `married`: A factor with levels `No` and `Yes` indicating whether the individual was married
-   `region`: A factor with levels `South`, `East`, and `West` indicating the region of the US the individual is from
-   `balance`: Average credit card balance in \$.

```{r}
#| warning: false
#| message: false

credit <- read_csv("data/credit.csv") |>
  mutate(maxed = if_else(balance == 0, 1, 0),
         student = as.factor(student))
```
# Part 1: Linear Regression

**The objective of this analysis is to predict whether a person average card balance.**

## Exercise 1

Consider the following models:

```{r}
model1 <- lm(balance ~ income + limit + age + rating, data = credit)


vif(model1)
```

What does VIF stand for? How do you use it? What conclusions can you draw from the output above.

## Exercise 2

```{r}
model2 <- lm(balance ~ income + limit + student + limit*student, data = credit)

tidy(model2) |>  kable()
```

Describe and interpret the interaction term from the above model. Be sure to give the value of the coefficient an describe what it means.

## Exercise 3

Consider the following analysis:

```{r}
model3 <- lm(balance ~ income + limit + age + rating + limit*student, data = credit)

anova(model2, model3)
```

Describe the test above. Write down the null and alternative hypotheses, test statistic, p-value, and interpret the results.

## Exercise 4

Based on the previous question and the output below, which model is better? Cite at least three forms of evidence.

```{r}
glance(model2) |> kable()
glance(model3) |> kable()
```



# Part 2: Logistic Regression

**The objective of this analysis is to predict whether a person has maxed out their credit card, i.e., had \$0 average card balance.**

## Exercise 1

-   Why is logistic regression the best modeling approach for this analysis?

## Exercise 2

```{r}
#| echo: FALSE

credit |> 
  count(maxed)
```

Let $\pi$ represent the probability that a person has maxed out their credit card.

-   Compute the empirical probability
-   Compute the empirical odds
-   Compute the empirical log odds

## Exercise 3

Consider the following code and output:

```{r}
credit_rec <- glm(maxed ~ income + age + student, 
                     data = credit, family = "binomial")
credit_rec |> tidy() |> kable()
```

-   Write out the formula for the probability of "success" that corresponds to this model.
-   Write out the formula for the log-odds of "success" that corresponds to this model.



## Exercise 4

What condition of logistic regression can be assessed with this plot. Does it appear that that condition is satisfied?

```{r}

emplogitplot1(maxed ~ income, data = credit, ngroups = 10)
```

## Exercise 5

What is wrong with the procedure below?

```{r}
#| label: cv-model2

model1 <- glm(maxed ~ income + age + student, data = credit, family = "binomial")
model2 <- glm(maxed ~ income + rating + limit + age, data = credit, family = "binomial")

anova(model1, model2, test = "Chisq")
```



## Exercise 6

Consider `income` below. Interpret the coefficient of income. There are two different answers I will accept here. Once this is done, interpret the p-value associated with `income`. Make sure to:

-   State the null and alternative hypotheses
-   Identify the test statistic
-   State the distribution used to calculate the p-value
-   State the conclusion of the test at a significance level of $\alpha = 0.01$

```{r}
model2 |> tidy()
```



